---
title: "State Drug Utilization - Tobacco Cessation in Kentucky"
author: 'Michael Q. Maguire, MS'
date: 'March 7th, 2022'
always_allow_html: true
output: html_document
---
  
## Packages Used
  
```{r packageLoad, message = FALSE}
library(tidyverse)
library(data.table)
library(DT)
library(httr)
library(rvest)
library(hrbrthemes)
library(readxl)
```

```{r sdudFile, include = FALSE}
file_location <- 'C:/Users/michaelqmaguire2/Dropbox (UFL)/01_projects/medicaid-drug-utilization/medicaid-drug-utilization/data/clean/03_sdud-redbook-final.csv'
```

## Read in the SDUD data set

```{r sdudLoad}
sdud <- fread(file_location,
              colClasses = c('proper_ndc' = 'character'))
```

## Read in AG NDC file.

## Subset to Florida Only and Create MOUD Flags

```{r NDC}
ndc <-
  readxl::read_excel('./data/raw/Chantix varenicline NDC codes.xlsx')

ndc_patterns <- ndc |>
  tidyr::separate(
    'Chantix_Varenicline_NDC',
    into = c('code1', 'code2', 'code3'),
    sep = '-',
    remove = FALSE
  ) |>
  mutate(
    pattern = paste0(nchar(trimws(code1)), nchar(trimws(code2)), nchar(trimws(code3))),
    ndcnum = case_when(
      pattern == '532' ~ paste0(code1, paste0("0", code2), code3),
      pattern == '442' ~ paste0(paste0("0", code1), code2, code3),
      pattern == '443' ~ paste0(paste0("0", code1), code2, code3),
      pattern == '541' ~ paste0(code1, code2, paste0("0", code3)),
      TRUE ~ 'Pattern not matched'
    )
  )

ndc_chantix <- ndc_patterns$ndcnum
```

Variable `moudFlag` represents whether the brand name has a pattern match with the below values.
Variable `medication` represents the categorization according to AG.

```{r sdudFL}
sdudTobacco <- sdud[i = (data.table::between(year, 2012, 2021) &
                           proper_ndc %in% ndc_chantix),
                    j = .(
                      utilization.type,
                      state,
                      year,
                      quarter,
                      suppression,
                      proper_ndc,
                      gennme,
                      prodnme,
                      numberrx
                    )][,
                       `:=` (yearQuarter = zoo::as.yearqtr(format(paste0(year, quarter)), "%Y%q"))]

fwrite(sdudTobacco,
       file = paste0(
         './data/',
         format(Sys.Date(), '%Y%m%d'),
         "_all-states-chantix.csv"
       ))
```

The following NDC's were found in this data base:

```{r specificNDC}
received_ndc <- length(ndc_chantix)
matched_ndc <- length(unique(sdudTobacco$proper_ndc))

paste0("Proportion matched to SDUD: ",
       round(matched_ndc / received_ndc, 2))
```

## Aggregating overall numbers by State

```{r aggregateOverall}
sdudTobaccoAggState <- sdudTobacco[i = suppression == 'F',
                                   j = .(totalRX = sum(numberrx)),
                                   by = .(yearQuarter, state)]

setorder(sdudTobaccoAggState, state, yearQuarter)

fwrite(
  sdudTobaccoAggState,
  file = paste0(
    './data/',
    format(Sys.Date(), '%Y%m%d'),
    '_tobacco-cessation-national-aggregate.csv'
  )
)
```

```{r plot, fig.width = 18, fig.height = 10}
kyRawCessation <- sdudTobaccoAggState |> filter(state == 'KY') |>
  ggplot() +
  geom_col(aes(
    x = as.factor(yearQuarter),
    y = totalRX,
    fill = as.factor(yearQuarter)
  ), color = 'black') +
  ggtitle("State Drug Utilization Data - Tobacco Cessation Products in Kentucky, 2012 - 2021") +
  xlab('Year - Quarter') +
  ylab('Total Units Dispensed') +
  theme_ipsum_rc(grid = 'Y') +
  theme(
    axis.text.x = element_text(angle = 90, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    legend.position = 'none'
  ) +
  scale_fill_viridis_d()

kyRawCessation

jpeg(
  paste0(
    './plots/',
    format(Sys.Date(), '%Y%m%d'),
    "_ky-varenicline-raw.jpeg"
  ),
  width = 18,
  height = 10,
  units = 'in',
  res = 1200
)
kyRawCessation
dev.off()
```


## Other States

```{r, test}
noKy <- sdudTobaccoAggState[sdudTobaccoAggState$state != 'KY',]

noKyFormatted <- copy(noKy)

noKyFormatted <-
  noKyFormatted[,  `:=` (yearQtrF = as.character(yearQuarter), yearQuarter = NULL)][, j = .(state, yearQtrF, totalRX)]

fwrite(
  noKyFormatted,
  file = paste0(
    './data/',
    format(Sys.Date(), '%Y%m%d'),
    "_non-ky-medicaid-population-by-month.csv"
  )
)

datatable(noKyFormatted)
```

## Pull in Medicaid Enrollment numbers from CDC API

```{r, apiPull}
url <-
  'https://data.medicaid.gov/api/1/datastore/query/6c114b2c-cb83-559b-832f-4d8b06d6c1b9/0'

postResponse <- httr::POST(url)

postToChar <- rawToChar(postResponse$content)

charToJSON <- jsonlite::fromJSON(postToChar)

medicaidEnrollment <- as.data.table(charToJSON$results)

fwrite(
  x = medicaidEnrollment,
  file = paste0(
    './data/',
    format(Sys.Date(), '%Y%m%d'),
    "_non-ky-medicaid-population-by-month.csv"
  )
)
```

```{r, kyFromAPI}
medicaidEnrollmentKY <- medicaidEnrollment[i = (state == 'Kentucky')]

medicaidEnrollmentKYQtr <- copy(medicaidEnrollmentKY)

medicaidEnrollmentKYQtr <-
  medicaidEnrollmentKYQtr[, `:=` (
    quarter = fcase(
      enrollment_month %in% c('1', '2', '3'),
      1,
      enrollment_month %in% c('4', '5', '6'),
      2,
      enrollment_month %in% c('7', '8', '9'),
      3,
      enrollment_month %in% c('10', '11', '12'),
      4,
      default = 5
    )
  )][,
     `:=` (
       yearQuarter = zoo::as.yearqtr(format(paste0(
         enrollment_year, quarter
       )), "%Y%q"),
       totalEnrollees = as.numeric(total_medicaid_enrollees)
     )]

setorder(medicaidEnrollmentKYQtr, yearQuarter)

medicaidEnrollmentKYQtr |> janitor::tabyl(enrollment_month, quarter)

medicaidEnrollmentKYQtr |> janitor::tabyl(yearQuarter, quarter)

fwrite(
  medicaidEnrollmentKYQtr,
  file = paste0(
    './data/',
    format(Sys.Date(), '%Y%m%d'),
    "_ky-medicaid-by-month-before-aggregation.csv"
  )
)
```

## Aggregate Medicaid Enrollment Numbers for KY

```{r, kyMedicaidAgg}
medicaidEnrollmentKyAgg <-
  medicaidEnrollmentKYQtr[, .(enrolleesPerQtr = sum(totalEnrollees) / 3), by = .(state, yearQuarter)]

setorder(medicaidEnrollmentKyAgg, yearQuarter)

sdudTobaccoAggStateKy <-
  sdudTobaccoAggState[i = (state == 'KY')][, `:=` (stateFull = 'Kentucky')]

setorder(sdudTobaccoAggStateKy, yearQuarter)

sdudWMedicaid <-
  sdudTobaccoAggStateKy[medicaidEnrollmentKyAgg, on = .(yearQuarter)]

sdudMedicaidFinal <-
  sdudWMedicaid[, `:=` (ratePer1k = (totalRX / enrolleesPerQtr) * 1000)]

fwrite(sdudMedicaidFinal,
       file = paste0(
         './data/',
         format(Sys.Date(), '%Y%m%d'),
         "_ky-varenicline-per-1k.csv"
       ))
```

## Plotting Medicaid Enrollees per Quarter in KY

```{r, medicaidPopAgg, fig.width = 18, fig.height = 10}
medicaidPopulationPlot <- sdudMedicaidFinal |> ggplot() +
  geom_col(aes(
    x = as.factor(yearQuarter),
    y = enrolleesPerQtr,
    fill = as.factor(yearQuarter)
  ), color = 'black') +
  scale_fill_viridis_d() +
  theme_ipsum_rc(grid = 'Y') +
  theme(
    axis.text.x = element_text(angle = 90, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    legend.position = 'none'
  ) +
  ggtitle("Total Medicaid Enrollees in KY, 2014 - 2020") +
  xlab("Year - Quarter") +
  ylab("Total Enrollees")

medicaidPopulationPlot

jpeg(
  paste0(
    './plots/',
    format(Sys.Date(), '%Y%m%d'),
    "_ky-medicaid-pop.jpeg"
  ),
  width = 18,
  height = 10,
  units = 'in',
  res = 1200
)
medicaidPopulationPlot
dev.off()
```

## Plotting Chantix Rate per 1k

```{r, chantixPer1k, fig.width = 18, fig.height = 10}
chantixPer1k <- sdudMedicaidFinal |> ggplot() +
  geom_col(aes(
    x = as.factor(yearQuarter),
    y = ratePer1k,
    fill = as.factor(yearQuarter)
  ), color = 'black') +
  scale_fill_viridis_d() +
  theme_ipsum_rc(grid = 'Y') +
  theme(
    axis.text.x = element_text(angle = 90, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    legend.position = 'none'
  ) +
  ggtitle("Varenicline Dispensed in Kentucky per 1,000 Medicaid Enrollees") +
  xlab("Year - Quarter") +
  ylab("Rate per 1,000 Enrollees")

chantixPer1k

jpeg(
  paste0(
    './plots/',
    format(Sys.Date(), '%Y%m%d'),
    "_ky-varenicline-per-1k.jpeg"
  ),
  width = 18,
  height = 10,
  units = 'in',
  res = 1200
)
chantixPer1k
dev.off()
```